Iterative Monomorphisation

### 2024-02-19 iterative_monomorphisation.tex
edits prelims

{\obeylines

\noindent\kern\parindent
  * algorithm works a priori with any rank-1 polymorphic logic

  * in particular, it works with polymorphic first- and higher-order logic as 
    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-201x,th1}

  * the algorithm does not need to inspect the precise structure of formulae
  * instead, it works on monomorphic or polymorphic instances of symbols that 
    occur in the formulae
  * type quantification is implicit, meaning that all type variables occurring
    in a formula are assumed to be universally quantified at the formula's top
    level
}

\bigskip

@@

\begin{definition}%[Type]
A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

@@

\begin{definition}%[Function symbol]
A (function or predicate) \emph{symbol} \(f\) has a type arity that defines the number of type arguments it takes. These arguments are listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type.
\end{definition}

{
\obeylines

* substitutions

* matching
}

@@

TODO: Use $\Phi_1$ etc.

>>>>

This algorithm works independently of the structure of the problem's formulae. It relies exclusively on the formulae's monomorphic and polymorphic symbol instances. Type variables are assumed to be universally quantified at a formula's top level. The precise form of formulae is left unspecified.
%
Due to this generality, iterative monomorphisation can be used with any reasonable rank-1 polymorphic logic. In particular, it can operate in the polymorphic first- and higher-order logics embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}, implemented by several automatic provers.

%The following terms are used in the algorithm descriptions.

%{\obeylines
%
%\noindent\kern\parindent
%  * algorithm works a priori with any rank-1 polymorphic logic
%
%  * in particular, it works with polymorphic first- and higher-order logic as 
%    embodied by TPTP's TF1 and TH1 syntaxes \cite{blanchette-paskevich-2013,th1}
%
%  * the algorithm does not need to inspect the precise structure of formulae
%  * instead, it works on monomorphic or polymorphic instances of symbols that 
%    occur in the formulae
%  * type quantification is implicit, meaning that all type variables occurring
%    in a formula are assumed to be universally quantified at the formula's top
%    level
%}

@@

Our abstract framework relies on the following basic definitions.

\begin{definition}%[Type]
A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

@@

\begin{definition}%[Function symbol]
A (function or predicate) {symbol} \(f\) has a \emph{type arity} that defines the number of type arguments it takes. A \emph{symbol instance} is a symbol applied to type arguments listed between angle brackets: \(f\langle \tau_1, \dots, \tau_n\rangle\), where each $\tau_i$ is a type. If $n = 0$, we omit the brackets (e.g.\ $f$).
\end{definition}

\begin{definition} % HEAVILY inspired from 2.4 in mono-trans (it is the definition of a substitution, there's bound to be some comonality between all of them) 
A (type) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
$\sigma = \{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$. They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
%The substitution %$\{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$
%is \emph{monomorphising} if the types $\tau_i$ are all monomorphic.
\end{definition}

\begin{definition}
Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\upsilon = \tau\).
\end{definition}

In the following sections, we will always match a polymorphic type $\upsilon$ against a monomorphic type $\tau$.
% and, on success, obtain a monomorphising substitution.

%{
%\obeylines
%
%* substitutions
%
%* matching
%}

@@

TODO: Use $\Phi$ for problem and $\varphi_1$ etc.
